---
title: "GDP Correlation Analysis"
---

The goal of this project is to examine the correlation between GDP per capita and three different variables: adult literacy rates, life expectancy, and measles immunisation rate. First I will run the correlation analysis for each variable. The data comes from the World Bank’s World Development Indicators [@worldbank2022]. Code is created with the help of GitCopilot [@copilot].

```{r}
setwd("/Users/Lisa/Desktop/QTM/assignment5")
library(readr)
library(ggplot2)
library(dplyr)

wdi <- read_csv("wdi.csv")

# Filter for rows with both gdp_per_capita and adult_literacy_rate
wdi_filtered <- subset(wdi, !is.na(gdp_per_capita) & !is.na(adult_literacy_rate))

# GDP and Adult Literacy Correlation
cor_literacy <- cor.test(
  wdi$gdp_per_capita, wdi$adult_literacy_rate,
  use = "complete.obs", method = "pearson"
)

# Filter for rows with both gdp_per_capita and Life Expectancy
wdi_filtered <- subset(wdi, !is.na(gdp_per_capita) & !is.na(life_expectancy))

# GDP and Life Expectancy
cor_life <- cor.test(
  wdi$gdp_per_capita, wdi$life_expectancy,
  use = "complete.obs", method = "pearson"
)

# Remove rows with missing GDP per capita or  measles immunisation rate data
wdi_filtered <- subset(wdi, !is.na(gdp_per_capita) & !is.na(measles_immunisation_rate))

# GDP and Measles Immunisation Rate
cor_immunisation <- cor.test(
  wdi$gdp_per_capita, wdi$measles_immunisation_rate,
  use = "complete.obs", method = "pearson"
)

# Create summary table
cor_table <- data.frame(
  Comparison = c(
    "GDP per Capita vs. Adult Literacy Rate",
    "GDP per Capita vs. Life Expectancy",
    "GDP per Capita vs. Measles Immunisation Rate"
  ),
  Correlation = c(
    round(cor_literacy$estimate, 3),
    round(cor_life$estimate, 3),
    round(cor_immunisation$estimate, 3)
  ),
  P_Value = c(
    signif(cor_literacy$p.value, 3),
    signif(cor_life$p.value, 3),
    signif(cor_immunisation$p.value, 3)
  )
)

# Display the table
knitr::kable(cor_table, caption = "Summary of Correlation Analyses")
```

#### Interpretation
All three variables are positivly correlated with GDP. Of all the variables, life expectancy has the strongest and most significant correlation with GDP (Figure 1). Adult literacy has the least significant, but second strongest correlation (Figure 2). Measles immunisation rate has the weakest correlation with GDP (Figure 3).

#### Figure 1
```{r echo=FALSE, fig.cap="Scatter plot of GDP per Capita vs. Life Expectancy"}
# Create the scatter plot
ggplot(wdi_filtered, aes(x = gdp_per_capita, y = life_expectancy)) +
    geom_point(alpha = 0.7) +
    labs(
        title = "GDP per Capita vs. Life Expectancy (2022)",
        x = "GDP per Capita (USD)",
        y = "Life Expectancy (Years)"
    ) +
    theme_minimal()
```

#### Figure 2
```{r echo=FALSE, fig.cap="Scatter plot of GDP per Capita vs. Adult Literacy Rate"}
# Create the scatter plot
ggplot(wdi_filtered, aes(x = gdp_per_capita, y = adult_literacy_rate)) +
    geom_point(alpha = 0.7) +
    labs(
        title = "GDP per Capita vs. Adult Literacy Rate (2022)",
        x = "GDP per Capita (USD)",
        y = "Adult Literacy Rate (%)"
    ) +
    theme_minimal()
```

#### Figure 3
```{r echo=FALSE, fig.cap="Bar plot of GDP per Capita and Measles Immunisation Rate"}
# Create GDP per capita groups
wdi_filtered <- wdi_filtered %>%
  mutate(gdp_group = factor(
    case_when(
      gdp_per_capita < 5000 ~ "Low",
      gdp_per_capita < 20000 ~ "Medium",
      TRUE ~ "High"
    ),
    levels = c("Low", "Medium", "High")
  ))

# Calculate average measles immunisation rate for each GDP group
avg_immunisation <- wdi_filtered %>%
  group_by(gdp_group) %>%
  summarise(mean_immunisation = mean(measles_immunisation_rate, na.rm = TRUE))

# Create the bar chart
ggplot(avg_immunisation, aes(x = gdp_group, y = mean_immunisation)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
    title = "Average Measles Immunisation Rate by GDP per Capita Group",
    x = "GDP per Capita Group",
    y = "Average Measles Immunisation Rate (%)"
  ) +
  theme_minimal()
```
