---
title: "GDP and Development Indicators"
author: "Lisa Blackmer-Raynolds"
date: "2025-06-04"
format:
  revealjs:
    theme: solarized
    transition: fade
---

# GDP and Development Indicators
### Lisa Blackmer-Raynolds  
### QTM 350  
### June 4, 2025

---

## Introduction

- Exploring the relationship between GDP per capita and development indicators
- Data: World Bank World Development Indicators (2022)

---

## Correlation Analysis

- GDP vs. Life Expectancy
- GDP vs. Adult Literacy Rate
- GDP vs. Measles Immunisation Rate

```{r echo=FALSE} 
library(readr)
library(ggplot2)
library(dplyr)

wdi <- read_csv("wdi.csv")

# Filter for rows with both gdp_per_capita and adult_literacy_rate
wdi_literacy <- subset(wdi, !is.na(gdp_per_capita) & !is.na(adult_literacy_rate))

# GDP and Adult Literacy Correlation
cor_literacy <- cor.test(
  wdi$gdp_per_capita, wdi$adult_literacy_rate,
  use = "complete.obs", method = "pearson"
)

# Filter for rows with both gdp_per_capita and Life Expectancy
wdi_life <- subset(wdi, !is.na(gdp_per_capita) & !is.na(life_expectancy))

# GDP and Life Expectancy
cor_life <- cor.test(
  wdi$gdp_per_capita, wdi$life_expectancy,
  use = "complete.obs", method = "pearson"
)

# Remove rows with missing GDP per capita or  measles immunisation rate data
wdi_measles <- subset(wdi, !is.na(gdp_per_capita) & !is.na(measles_immunisation_rate))

# GDP and Measles Immunisation Rate
cor_immunisation <- cor.test(
  wdi$gdp_per_capita, wdi$measles_immunisation_rate,
  use = "complete.obs", method = "pearson"
)

# Create summary table
cor_table <- data.frame(
  Comparison = c(
    "GDP per Capita vs. Adult Literacy Rate",
    "GDP per Capita vs. Life Expectancy",
    "GDP per Capita vs. Measles Immunisation Rate"
  ),
  Correlation = c(
    round(cor_literacy$estimate, 3),
    round(cor_life$estimate, 3),
    round(cor_immunisation$estimate, 3)
  ),
  P_Value = c(
    signif(cor_literacy$p.value, 3),
    signif(cor_life$p.value, 3),
    signif(cor_immunisation$p.value, 3)
  )
)

# Display the table
knitr::kable(cor_table, caption = "Summary of Correlation Analyses")
```

---

## Life expectancy has the strongest correlation with GDP
```{r echo=FALSE, fig.cap="Scatter plot of GDP per Capita vs. Life Expectancy"}
# Create the scatter plot
ggplot(wdi_life, aes(x = gdp_per_capita, y = life_expectancy)) +
    geom_point(alpha = 0.7) +
    labs(
        title = "GDP per Capita vs. Life Expectancy (2022)",
        x = "GDP per Capita (USD)",
        y = "Life Expectancy (Years)"
    ) +
    theme_minimal()
```

---
## Adult literacy is also positively correlated
```{r echo=FALSE, fig.cap="Scatter plot of GDP per Capita vs. Adult Literacy Rate"}
# Create the scatter plot
ggplot(wdi_literacy, aes(x = gdp_per_capita, y = adult_literacy_rate)) +
    geom_point(alpha = 0.7) +
    labs(
        title = "GDP per Capita vs. Adult Literacy Rate (2022)",
        x = "GDP per Capita (USD)",
        y = "Adult Literacy Rate (%)"
    ) +
    theme_minimal()
```

---
## Measles immunisation rate has the weakest correlation
```{r echo=FALSE, fig.cap="Bar plot of GDP per Capita and Measles Immunisation Rate"}
# Create GDP per capita groups
wdi_measles <- wdi_measles %>%
  mutate(gdp_group = factor(
    case_when(
      gdp_per_capita < 5000 ~ "Low",
      gdp_per_capita < 20000 ~ "Medium",
      TRUE ~ "High"
    ),
    levels = c("Low", "Medium", "High")
  ))

# Calculate average measles immunisation rate for each GDP group
avg_immunisation <- wdi_measles %>%
  group_by(gdp_group) %>%
  summarise(mean_immunisation = mean(measles_immunisation_rate, na.rm = TRUE))

# Create the bar chart
ggplot(avg_immunisation, aes(x = gdp_group, y = mean_immunisation)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
    title = "Average Measles Immunisation Rate by GDP per Capita Group",
    x = "GDP per Capita Group",
    y = "Average Measles Immunisation Rate (%)"
  ) +
  theme_minimal()
```

---
## Conclusion

- GDP per capita is strongly associated with life expectancy and adult literacy rate.
- The relationship with measles immunisation rate is weaker, suggesting other factors may play a role.